---
title: "RBootcamp 2017 -- Getting Started"
output: 
  slidy_presentation:
    duration: 45
    incremental: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
```

## Getting Started: Agenda

- RStudio setup

- Basics review and more

- Saving and sourcing scripts

- Installing and using packages

## RStudio setup

- Start RStudio

- Go ahead, start it up

    - We're done

    - Just kidding

- Check out the four panes 

    - Do you only see three panes?
    
        - Prob b/c no files to view yet in the top-left pane

    - Each pane has its special set of tabs
    
- As we go through the material this week we will demonstrate useful
aspects of each pane

## Exercise 1
    
1.1 In the Console pane give x the value one  `x <- 1`
    
    - What changed in the upper left pane?
    
1.2 Use the History tab to re-execute the same command
    
## RStudio advanced environment

- The Menu bar

    - I use RStudio icons more than the menu items

    - the Tools menu (~ Settings)

        - Try changing the font
    
        - Try rearranging the panes or the underlying tabs
        
        - Check out which version of R is being used

    - the Files menu
    
        - I use "Save As"
        
    - the Help menu
    
        - where the cheatsheets are!

    - the Session menu
    
        - Restart R
        
- As we go through the material this week we will demonstrate the usefulness of RStudio's advanced settings

## Basics review and more

>- Do some basic math in the console window

```{r, echo=TRUE}
2 + 5
-2^0.5 # which operator takes precedence
sqrt(2)
abs(-15.4)
3 > 4 # logical
```

## Syntax

The table below can be obtained by asking for help on "Syntax"  
`?Syntax` or `help("Syntax")`

Symbol | Meaning
-------|--------
--------------- | ---------------------------------------
* /	|multiply, divide
+ -	|(binary) add, subtract
^	|exponentiation (right to left)
- +	|unary minus and plus
--------------- | ---------------------------------------
< > <= >= == !=	|ordering and comparison
!	|negation
& &&	|and
| ||	|or
--------------- | ---------------------------------------
<- <<-	|assignment (right to left)
=	|assignment (right to left)
-> ->>	|rightwards assignment (rarely used)
--------------- | ---------------------------------------
:|sequence operator
%any%	|special operators (including %% and %/%)
~	|as in formulae
--------------- | ---------------------------------------
:: :::	|access variables in a namespace
$ @	|component / slot extraction
[ [[	|indexing
--------------- | ---------------------------------------
?	|help (unary and binary)

## What data types can R work with?

Type | Example  | Other languages
-----|--------|------------------
Logical | TRUE, FALSE | Logical
Integer | 1L, -2L (rarely important to use 'L') | Integer
Numeric | 1.214, 1, -2 | double, real
Character | 'a', "B", "Now is the time ..." | strings
function  | sqrt  | function, subroutine

- More similar to FORTRAN than c

- More complex structures (vectors, lists, ...) will be discussed later this morning

##  Exercise 2

2.1 In the Console, store the value 6 in the variable `r` in one line
then in the next line calculate the area of a circle with radius `r`
(hint: the symbol in R for pi is `pi`)

2.2 What is the symbol in R for the largest numeric possible? Smallest?

2.3 What do you get when you divide `1` by `0`? `0` by `0`?

## Not available

- R goes out of its way to record a missing value for a variable by using
the special value `NA`

    - This is a major strength of R

    - (R uses the value `NaN` for "not a number", but that's different from `NA`)

- Testing for and detecting missing data is **an extremely important step** in data cleansing and analysis

    - `?is.na`

- Some functions ignore NA's, others do not

- The existence of NA's in your data will be a frequent source of trip-ups
in your data analysis

- Examples are better explored after vectors are introduced

## The working directory

>- **Where are my files?**
is a frequently asked question in R / RStudio

```{r eval=TRUE, echo=TRUE, size='tiny'}
getwd()
```

>- How do I get somewhere else?

```{r eval=FALSE, results='hide', size='tiny'}
setwd("Desktop")
```

>- ... or use the Session menu ...

>- ... or start in the desired directory by double-clicking
an appropriate file and let RStudio set the working directory

>- Try to stick with relative pathnames to make your work portable
from machine to machine to cloud to company/client network to ...

## Directory paths

- Windows paths are constructed with the backward slash "\\" separator

- mac, *nix machines construct paths with the forward slash "/" separator

- Furthermore, in R, the backward slash "\\" 
is used for things like tabs and newline characters

    - If you really need a slash and aren't intending to "escape" the next character, just type it twice

- Foreshadowing: For a painless, OS-independent approach to constructing
paths to files, 
use the `file.path` function

## Script files

- Scripts are text files that store R code

- **Extremely** important

- You will use scripts, a.k.a. "source files", frequently

- They will store the programs you write

## Exercise 3

3.1 Create a new script that stores the two lines from Exercise 2.1, area of a circle

3.2 Run the script line-by-line.

3.3 Run the script in its entirety by "Source-ing" it with the 
"Source" icon above the code pane

>- What was the difference?

3.4 Save the script with the name Test.R

>- How do you know it was saved?

3.5 In the Console pane, "source" the file from it's stored location using its full name

## Installing and using packages

- An R "package" is a collection of R code and objects that,
in the eyes of the author(s),
facilitate
some computation, analysis, etc.

- R's package facility is arguably what makes the language so powerful

- To use a package

    - First, download it and install it

        - The easiest way to download and install is with RStudio's "Packages" menu
    
        - Packages are installed in the library folder of the version of R you are running
    
    - Second, 
    
        - either load the entirety of its capabilities with the `library` function
        
        - or use a specific object from the package using the construct
        `packageName::objectName`
        
        - You will almost always use the `library` function
        
- You already installed the `raw` package and its suggested packages

- We will be using many packages in this camp,
so we will not belabor the point now and move along
        
## Exercise 4

4.1 Install the `actuar` package

4.2 Get help on the package contents (hint: `help(package = actuar)`)

4.3 Get help on the package's 
**Empirical Limited Expected Value** function (`elev`)

>- Did you get the message

`No documentation for 'elev' in specified packages and libraries:`  
`you could try '??elev'`