---
title: "Data Structures"
output:
  slidy_presentation:
    duration: 45
    fig_height: 3
    fig_width: 6
  beamer_presentation:
    fonttheme: professionalfonts
---



## Data Structures in R

Most of the data you will be working with in R will be one of the following

  + data.frame
  + vector
  + list
  
You will also see some matrices and factors, but let's skip those for now.



## Pre-reading

You guys all did the pre-reading, right?



## Data Frames: The Short Story

1. A data.frame is a list, too
2. Unlike other lists, all elements have to be vectors
3. All vectors in a data.frame have to be the same length
4. A data.frame is like a SQL table or a table in Excel, so you will work with them a lot



## Okay, so what is a list, then?

1. A list is a bag of stuff
2. Everything in the list has an index (1, 2, 3, etc.)
3. The stuff can be anything

  * vectors
  * other lists
  * special objects
  * etc.

  

## Vectors: a Novella

1. Vectors are indexed, like lists
2. They can only store one type of data (logical, integer, numeric, character)
3. At each index, you can only have a single value



## Review

1. lists are bags of stuff
2. vectors are sequences of single values (all of the same type)
3. data.frames are lists made up of vectors (all of the same length)



## Get Your Fingers Moving
```{r eval=TRUE}
w <- 6
x <- 3L
y <- TRUE
z <- 'finagle a bagel'
df1 <- iris
l1 <- list(w, x, y, z, df1)
```



## A Moment for Introspection
Use the following functions to learn stuff about your data

* typeof
* class
* mode
* str
* is.integer
* is.numeric
* is.logical
* is.character
* is.double


## Observations?




## Observations
1. mode is mostly meaningless
2. For vectors, class and typeof kind of tell the same thing
3. is.* functions are good for vectors
4. str tells something good for everything



## `seq`

`seq` is used often to generate a sequence of values. The colon operator `:` is a shortcut for a sequence of integers.

```{r }
pies = seq(from = 0, by = pi, length.out = 5)
i <- 1:5
year = 2000:2004
```


## `rep`

The `rep` function will replicate its input 
```{r }
i = rep(pi, 100)
head(i)
```


## Concatenation

The `c()` function will concatenate values for lists and vectors.

`cbind` concatenates columns.

`rbind` concatenates rows.



```{r results='hide'}
i <- c(1, 2, 3, 4, 5)
j <- c(6, 7, 8, 9, 10)
k <- c(i, j)
l <- c(1:5, 6:10)
```




## Vector access - by index

Vectors may be accessed by their numeric indices. Remember, ':' is shorthand to generate a sequence.

```{r }
e <- rnorm(100)
e[1]
e[1:4]
e[c(1,3)]
```




## Vector access - logical access

Vectors may be accessed logically. This may be done by passing in a logical vector, or a logical expression.

```{r }
i = 5:9
i[c(TRUE, FALSE, FALSE, FALSE, TRUE)]
i[i > 7]
b = i > 7
b
i[b]
```




## `order` and `sort`

The function `order` will return the indices of the vector in order.

```{r }
x <- sample(letters)
x
order(x)
sort(x)
```



## Vector arithmetic

Vectors may be used in arithmetic operations. 

```{r eval=FALSE}
B0 <- 5
B1 <- 1.5

e <- rnorm(N, mean = 0, sd = 1)
X1 <- rep(seq(1,10),10)

Y <- B0 + B1 * X1 + e
```



## Recycling

R will "recycle" vectors until there are enough elements to perform an operation. Everything gets as "long" as the longest vector in the operation. For scalar operations on a vector this doesn't involve any drama. Try the following code:

```{r size='tiny'}
vector1 = 1:10
vector2 = 1:5
scalar = 3

print(vector1 + scalar)
print(vector2 + scalar)
print(vector1 + vector2)
```



## Set theory - Part I

The `%in%` operator will return a logical vector indicating whether or not an element of the first set is contained in the second set.

```{r }
x <- 1:10
y <- 5:15
x %in% y
```



## Set theory - Part II

* `union`
* `intersect`
* `setdiff`
* `setequal`
* `is.element`

```{r eval = FALSE}
?union
```

```{r }
x <- 1900:1910
y <- 1905:1915
intersect(x, y)
setdiff(x, y)
setequal(x, y)
is.element(1941, y)
```



## Summarization

Loads of functions take vector input and return scalar output. Translation of a large sest of numbers into a few, informative values is one of the cornerstones of statistics.

```{r eval=FALSE}
x = 1:50
sum(x)
mean(x)
max(x)
length(x)
var(x)
```




## Lists


```{r }
x <- list()
typeof(x)
x[[1]] <- c("Hello", "there", "this", "is", "a", "list")
x[[2]] <- c(pi, exp(1))
summary(x)
str(x)
```



## List names

Names are handled with the special `$` operator. `$` permits access to a single element. (A single element of a list can be a vector!)

```{r}
y <- list()
y[[1]] <- c("Lou Reed", "Patti Smith")
y[[2]] <- c(45, 63)

names(y) <- c("Artist", "Age")

y$Artist
y$Age
```



## Data Frames

```{r}
teams  <- c("Toronto", "New York", "Baltimore", "Tampa Bay", "Boston")
wins   <- c(93, 87, 81, 80, 78)
losses <- c(69, 75, 81, 82, 84)

alEast <- data.frame(Team = teams, W = wins, L = losses)
```



## Selecting rows and columns

```{r, eval=FALSE}
alEast[1, ]
alEast[, 2]
alEast[, c('W', 'L')]
alEast[3] 
alEast[c(T, T, F, F, F), ]
alEast[alEast$Team == 'New York', ]
```

What happens on the last one if you leave off the comma?



## Exploring data frames

```{r}
head(alEast)
tail(alEast)
names(alEast)
str(alEast)
```


## Summary

```{r}
summary(alEast)
```



