---
title: "Advanced Visualization"
author: "Brian A. Fannin"
date: "August 21, 2017"
output:
  tufte::tufte_handout: default
  revealjs::revealjs_presentation:
    center: no
    css: ./css/revealOpts.css
    reveal_plugins:
    - notes
    - zoom
    self_contained: no
    theme: solarized
    transition: slide
---

```{r echo=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))
knitr::opts_chunk$set(
    warning=FALSE
  , error=FALSE
  , echo=TRUE
  , message=FALSE
  , fig.height = 4.5
  , fig.pos="t"
  , collapse = TRUE
)
knitr::opts_knit$set(root.dir = normalizePath('../../'))
```

# Overview

## ggplot2

ggplot2 developed by Hadley Wickham, based on the "grammar of graphics". Particularly well suited (IMHO) for multi-dimensional, multivariate analysis.

Requires 3 things:

1. Data
2. Mapping
2. Geometric layers

## Data

```{r }
library(raw)
data(RegionExperience)
library(ggplot2)

basePlot <- ggplot(RegionExperience)
class(basePlot)
```

<div class="notes">
Notice that we assigned the result of the function call to an object called `basePlot`. This means we don't automatically get output. Take a bit of time to have a look at what's contained in the `basePlot` object. 
</div>

## Aesthetics

Aesthetics are anything visible on the plot. When an aesthetic is "mapped", it means that the data will define how the aesthetic appears. We'll map our aesthetics with the `aes` function.

```{r }
basePlot <- basePlot + aes(x = PolicyYear, y = NumClaims, color=Region)
```

## Adding layers

The `geom_*` functions add geometric shapes. 

```{r }
p <- basePlot + geom_line()
p
```

## Adding layers

```{r }
p <- basePlot + geom_point()
p
```

## Nothing wrong with adding two layers

```{r }
p <- basePlot + geom_point() + geom_line()
p
```

## One step

Typically we don't do this in steps.

```{r }
p <- ggplot(RegionExperience, aes(x = PolicyYear, y = NumClaims, group=Region, color=Region)) + geom_line()
p
```

## Layers have loads of parameters

```{r }
p <- basePlot + geom_bar(stat="identity", aes(fill = Region))
p
```

## Mmm, parameters

```{r }
p <- basePlot + geom_bar(stat="identity", position="dodge", aes(fill=Region))
p
```

## Layer parameters

How do I know which parameters do what? RTFM

Help, google and stack overflow are your friends. Also, loads of books on the subject.

# More features

## Facets

```{r }
data(StateExperience)
p <- ggplot(StateExperience, aes(x = PolicyYear, y = NumClaims, color = State)) + geom_point() + facet_wrap(~ Region)
p <- p + theme(legend.position = "none")
p
```

## Statistical smoothers

```{r}
p <- ggplot(RegionExperience, aes(x = PolicyYear, y = NumClaims, group=Region, color=Region)) + geom_point()
p + geom_smooth(se = FALSE)
```

## Linear smoother

```{r }
p + geom_smooth(method = lm)
```

# Scales

## Scales

Scales control how things render on the plot. We must scale our numbers to the plotting device (typically a section of a computer screen.) We can also map things like color and axes to data values and control formatting.

## Scales

```{r}
p + scale_y_continuous(labels = scales::comma)
```

## Scales

The use of scales is a _very_ detailed topic, particularly when we start talking about color. Your specific problem will likely require a bit of research and experimentation. StackOverflow.com is your friend.

## Non-data visual elements

Non-data elements are things like labels. Here's a sample of a few:

* `xlab()`, `ylab()` -> `plt + ylab("This is my y label") + xlab("Here is an x label")`
* `ggtitle()` ->  `plt + ggtitle("Title of my plot")`
* `labs()` -> `plt + labs(x = "x-axis title", title = "My title")`
* `theme_bw()`, `theme_minimal()` -> `plt + theme_bw()`
* The `theme()` function gives complete control over all non-data related visual elements
* Check out the `ggthemes` package

# Summary

## Summary

* `ggplot2` is difficult at first, but will repay your investment.
* Works very well with grouped data to color/facet points.
* Fine-tuning things like axis labels can be a headache, but will get easier. Yes, Excel makes it easier to add data labels and change colors. `ggplot2` makes it easier to work with data.

## Questions to ask when debugging

1. How am I mapping my data to an aesthetic?
2. How are these aesthetics handled by the geom that I'm using?
3. 

## Reference

* http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
* R Graphics Cookbook by Winston Chang
* http://vita.had.co.nz/papers/layered-grammar.html

## Your turn

1. Create a scatter plot for policy year and number of claims
2. Color each point based on region
2. Add a linear smoother. Which region is showing the greatest increase in claims?
4. Form the policy frequency by taking the ratio of claims to policies. Plot this.

Extra credit:

1. Use the state data to create a time series number of claims. Facet by region.

## Answers

```{r}
library(raw)
data("RegionExperience")
plt1 <- ggplot(RegionExperience, aes(x = PolicyYear, y = NumClaims)) + geom_point()
plt1

plt2 <- plt1 + aes(color = Region)
plt2

plt3 <- plt2 + stat_smooth(method = "lm")
plt3

RegionExperience$Frequency <- with(RegionExperience, NumClaims / NumPolicies)

plt4 <- ggplot(RegionExperience, aes(x = PolicyYear, y = Frequency, color = Region)) + geom_point() + geom_line() + stat_smooth(method = lm)
plt4
```

```{r}
data("StateExperience")
pltExtra <- ggplot(StateExperience, aes(x = PolicyYear, y = NumClaims, color = Postal)) + geom_point() + geom_line()
pltExtra + facet_wrap(~ Region)
```
