---
title: "Day One Review"
output:
  slidy_presentation:
    duration: 45
    fig_height: 3
    fig_width: 6
  beamer_presentation:
    fonttheme: professionalfonts
---



## Assignment
What is the assigment operator in R?



## Equality
What is the difference between "=" and "=="?
What operator do you use to test inequality?



## Let's have a little chat
* There are always a dozen ways to do something in R.
* Most of the time this is great, except for when you are brand new...
* So, just ignore that there are exceptions to everything we tell you. :(



## Our friend "="
* The assigment operator in R is "<-"
* When you are calling a function, sometimes you have to use "=", like
```{r}
1:5
seq(1, 5)
seq(from = 1, to = 5)
seq(to = 5, from = 1)
seq(1, 5, by = 2)
```




## Exercise
* Create a vector of years from 2007 to 2016
* Create a vector of random normal values of the same length
* Create a data.frame object that combines the two vectors together
* Compare your result to your neighbor
* Are they the same?  Why or why not?



## Let's go spelunking!
* Functions you can use to explore your junk
    + `str`
    + `class`
    + `ls`
    + `search`
    + `print` (or just type the junk's name)
* Are there any others?  



## Big C, Little c
* So what is the difference between big C and little c?



## R, you naughty thing
* Yes, R is case-sensitive.
* So, not naughty, just sensitive.
* Oh, and dots are special.
* "And don't forget me, the underscore!"



## [] versus ()
* Pick your favorite function (it's OK, this can be a short relationship)
* What do you get when you run it with `[]` afterward?
* With `()`
* With nothing? (scandal!)



## Plotting...to leave great reviews of this course
* High level plot functions in R
    + plot
    + curve
    + pairs
    + boxplot
    + barplot
    + hist
* Use the googler to try and find others



## Which is better?
* `ggplot2` or `base` plotting?



## Trick Question
For a review of `ggplot2` please see the cheatsheet on the community page



## Data import and export
* Write RData files using `save`
* Read RData files using `load`
* What functions do you need to read and write CSV files?
* What are some of the naughty things you need to worry about with CSV?


## Excel?
* Use every tool in the box
* But, just be careful with dates, factors, commas in numbers, etc.



## Questions

* Plot a lognormal distribution with a mean of $10,000 and a CV of 30%.
* For that distribution, what is the probability of seeing a claim greater than $100,000?
* Generate 100 and 1,000 observations from that distribution.
* Draw a histogram for each sample.
* What are the mean, standard deviation and CV of each sample?



## Answers
```{r }
severity <- 10000
CV <- .3
sigma <- sqrt(log(1 + CV^2))
mu <- log(severity) - sigma^2/2
plot(function(x) dlnorm(x), mu, sigma, ylab="LN f(x)")
```



## Hello Teacher
Psst, Adam.  You didn't cover X, Y or Z.  





## Day Two
What are we going to learn today?  A lot more valuable stuff than yesterday, that's for sure.

* dplyr
* RMarkdown
* ChainLadder

