---
title: "ChainLadder Package"
output: 
  slidy_presentation:
    duration: 45
  beamer_presentation:
    fonttheme: professionalfonts
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Agenda

Investigate stochastic methods in 
the ChainLadder package that are useful for reserving

Learn how to work with the objects returned by the methods

## Data triangles

ABC : WC

GenIns: Unknown line

RAA:  reinsurance GL data

auto: auto

Mortgage: mortgage guarantee

USAApaid, USAAincurred: CAS website

## Useful stochastic methods for reserving

- MackChainLadder

- Dave Clark's methods ("LDF Curve-Fitting ...",
CAS *eForum*)

    - ClarkLDF
    
    - ClarkCapeCod
    
- Other stochastic methods include ...

    - BootChainLadder: Bootstrapping
    
    - MunichChainLadder: paid & incurred simultaneously

    - tweedieReserve: glm including one-year risk horizon (Solvency II)
    
## MackChainLadder

- Develop the GenIns triangle to ultimate with an estimated tail factor

- How to find useful values in the object returned

    - Hint: learn what is returned by the **summary** function

```{r}
library(ChainLadder)
GenIns
M <- MackChainLadder(GenIns, tail = TRUE)
```

- Investigate items returned by 'summary'

```{r}
print(summary(M))
class(summary(M))
names(summary(M))
class(summary(M)$ByOrigin)
# Here is where the accident year ultimates are
summary(M)$ByOrigin$Ultimate
  # Aside: Using '$' sign same as pulling contents of "Ultimate" sublist
  summary(M)$ByOrigin[["Ultimate"]]
  # Contrast with the "Ultimate" sublist itself
  summary(M)$ByOrigin["Ultimate"]

# Back to business
class(summary(M)$Totals)
# Here is where total IBNR is
summary(M)$Totals["IBNR:", "Totals"]
summary(M)$Totals[4, 1]
# Here is the standard error of total IBNR
summary(M)$Totals["Mack S.E.:", "Totals"]
```

- You can fit a two parameter distribution
(e.g., lognormal)
to total IBNR using the estimated mean and standard error

## Clark's methods

- LDF Method: Develop the GenIns triangle to ultimate

    - Fits a development curve, therefore a tail factor implicitly

- Pull estimated IBNR and its standard error

```{r}
DC <- ClarkLDF(GenIns)
print(summary(DC))
class(summary(DC))
# Here is where the accident year ultimates are
summary(DC)[1:10, "UltimateValue"]
# Here is where total IBNR is
summary(DC)["Total", "FutureValue"]
# Here is the standard error of total IBNR
summary(DC)["Total", "StdError"]
```

- Fit a distribution to the mean and standard error

## Clark's methods

- CapeCod Method

    - Like the BF method but ELR is estimated, not provided
  
    - Needs Premium argument

- Pull estimated IBNR and its standard error

```{r}
# Premium sequence comes from Clark's paper
CC <- ClarkCapeCod(GenIns, Premium = 
                     seq(10e06, length = 10, by = 400000))
print(summary(CC))
class(summary(CC))
# Here is where the accident year ultimates are
summary(CC)[1:10, "UltimateValue"]
# Here is where total IBNR is
summary(CC)["Total", "FutureValue"]
# Here is the standard error of total IBNR
summary(CC)["Total", "StdError"]
```

- Fit a distribution to the mean and standard error

## Fit a lognormal distribution to mean and sd

```{r}
lnormParms <- function(mean, sd) {
  sigma2 <- log(1 + (sd / mean)^2)
  meanlog <- log(mean) - .5 * sigma2
  cbind(meanlog, sdlog = sqrt(sigma2))[,]
}
```

- The lognormal parameters matching the MackChainLadder estimate
of Total IBNR

```{r}
parms <- lnormParms(summary(M)$Totals["IBNR:", "Totals"],
           summary(M)$Totals["Mack S.E.:", "Totals"])
parms
```

- The 95th percentile of estimated IBNR is

```{r}
qlnorm(.95, parms[1], parms[2])
```

- What quantile corresponds to the mean

```{r}
plnorm(summary(M)$Totals["IBNR:", "Totals"], parms[1], parms[2])
```

- Not very skewed. What is the cv?

```{r}
summary(M)$Totals["Mack S.E.:", "Totals"] / 
  summary(M)$Totals["IBNR:", "Totals"]
```

- How compares with the CV from the ClarkLDF method?

```{r}
summary(DC)["Total", "StdError"] / 
  summary(DC)["Total", "FutureValue"]
# Also found in the object
summary(DC)["Total", "CV"]
```

