---
title: "dplyr"
author: "Brian A. Fannin"
date: "August 22, 2017"
output:
  revealjs::revealjs_presentation:
    center: no
    css: ./css/revealOpts.css
    reveal_plugins:
    - notes
    - zoom
    self_contained: no
    theme: solarized
    transition: slide
  tufte::tufte_handout: default
bibliography: bibliography.bib
---

```{r include=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))
knitr::opts_chunk$set(
    warning=FALSE
  , error=FALSE
  , echo=TRUE
  , message=FALSE
  , fig.height = 4.5
  , collapse = TRUE
)
knitr::opts_knit$set(root.dir = normalizePath('../../'))
```

# Overview

## 

Another element of the "tidyverse" developed by Hadley Wickham and many others.

## Let's look at some data

```{r}
library(raw)
data("MultiTri")
```

```{r eval=FALSE}
View(MultiTri)
```

## Real quick:

* What's the average IBNR for lag 2 by accident year?

## dplyr to the rescue!

__Basic verbs__

| Verb              |
|--------------------|
| select/rename |
| mutate |
| group  |
| arrange |
| filter |
| summarise |

# select/rename

Comparable to the "select" verb in SQL. This will return only those columns you've requested. A "negative" column name will omit that column.

```{r}
financials <- select(MultiTri, CumulativePaid, CumulativeIncurred, IBNR)
no_lag <- select(MultiTri, -Lag)
```

## Identifying which columns to select

Loads of helper functions to find columns

| function   |
|------------|
| contains  |
| starts_with |
| ends_with |
| matches   |
| num_range |

```{r}
years <- select(MultiTri, contains("year"))
```

## rename

Closely related to `select`. New name on the left, old name on the right. (Think the way that variables are usually assigned.)

```{r}
new_tri <- rename(MultiTri, DevelopmentLag = Lag)
```

# mutate

## mutate

Create a new column or alter an existing one.

```{r}
new_tri <- mutate(
  new_tri
  , PaidToIncurred = CumulativePaid / CumulativeIncurred
  , Upper = DevelopmentYear == 1997)
```

## Combining operations

```{r}
new_tri <- mutate(MultiTri, Upper = DevelopmentYear == 1997)
new_tri <- select(new_tri, -DevelopmentYear)
```

This syntax will get tedious real fast. We need some way to make this more efficient. Fortunately, we have one ...

# The pipe operator

## %>%

Key to getting the most out of dplyr is using the `%>%` or "pipe" operator. It takes whatever is to its left and inserts it as the first (unnamed) argument in the function to its left. 

Comes from its own package, "magrittr" and may be used without dplyr. 

In RStudio, may be inserted with CTRL-SHIFT-M

## Pipe example

```{r}
1 %>% exp()
```

## Chain

May be chained as often as you like. Operations happen left to right. The output keeps getting passed as the input of the next function:

```{r}
1 %>% exp() %>% log()
```

# arrange

## arrange

Use the desc function to arrange in descending order.

```{r eval=FALSE}
MultiTri %>% 
  arrange(AccidentYear)

MultiTri %>% 
  arrange(desc(IBNR))
```

# filter/slice

## Straightforward

```{r}
upper_tri <- MultiTri %>% 
  filter(DevelopmentYear <= 1997)
```

Multiple conditions are OK

```{r}
upper_tri <- MultiTri %>% 
  filter(DevelopmentYear <= 1997
         , IBNR > 500)
```

## slice

`slice` will take specific rows of data.

```{r}
every_fifth <- MultiTri %>%
  slice(seq(from = 5, by = 5, to = nrow(MultiTri)))
```

# group_by

## group_by

This will group the data. The effect isn't material until another operation is applied.

```{r}
df_grouped <- MultiTri %>% 
  group_by(Company, AccidentYear)
```


# summarise

## Apply a function across each group

```{r}
dfBigYear <- MultiTri %>% 
  group_by(AccidentYear) %>% 
  summarise(BiggestIBNR = max(IBNR))
```

## Gotcha

Arrange does not respect grouping! Didn't used to be this way. A rare misstep (IMHO) for Mr. Wickham.

# All together now!

## Answer a complex question quickly

* For each company which has had paid to incurred ratio less than 40%, which accident year had the highest P2I?

##

```{r}
dfBigCase <- MultiTri %>% 
  mutate(PaidToIncurred = CumulativePaid / CumulativeIncurred) %>% 
  filter(PaidToIncurred < 0.4) %>% 
  group_by(Company) %>% 
  arrange(desc(PaidToIncurred)) %>% 
  slice(1) %>% 
  select(Company, AccidentYear)
```


# Summary

## References

