---
title: "More dplyr practice"
author: "Brian A. Fannin"
date: "August 23, 2017"
output:
  revealjs::revealjs_presentation:
    center: no
    css: ./css/revealOpts.css
    reveal_plugins:
    - notes
    - zoom
    self_contained: no
    theme: solarized
    transition: slide
---

```{r include=FALSE}
knitr::opts_chunk$set(
    warning=FALSE
  , error=FALSE
  , echo=TRUE
  , message=FALSE
)
knitr::opts_knit$set(root.dir = normalizePath('../../'))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
```

# Yay, more dplyr!

## The Lahman data

Baseball stats

```{r }
library(Lahman)
```

```{r echo=TRUE, eval=FALSE}
library(Lahman)
data(package = 'Lahman')
```

## Your questions

1. What is mean and median salary by year?
1. Using the "Master" and "Salaries" tables, what is the min, max and mean salaries for all players who are (according to this data) alive. 
2. What is the ratio of the mean salary by year of left-handed batters to right-handed batters? Note: there are three kinds of batters in the database: "L", "R" and "B". Ignore ambidextrous (i.e. "B") batters.

## Answers - 1

```{r}
library(tidyr)
library(dplyr)
```

```{r}
dfSalaryByYear <- Salaries %>% 
  group_by(yearID) %>% 
  summarise(AvgSalary = mean(salary)
            , MedianSalary = median(salary))
```

## Answers - 2

```{r}
dfPlayerSalary <- Master %>% 
  filter(is.na(deathDate)) %>% 
  select(playerID, bats) %>% 
  inner_join(Salaries)

dfSalarySummary <- dfPlayerSalary %>% 
  summarise(MinSalary = min(salary)
            , MaxSalary = max(salary)
            , AvgSalary = mean(salary))
```

## Answers - 3

```{r}
dfYearAndHand <- dfPlayerSalary %>% 
  filter(bats != 'B') %>% 
  group_by(yearID, bats) %>% 
  summarise(AvgSalary = mean(salary)) %>% 
  spread(bats, AvgSalary) %>% 
  mutate(LeftRelativity = L / R)
```

