---
title: "Decision Trees"
output:
  slidy_presentation:
    duration: 45
    fig_height: 3
    fig_width: 6
  beamer_presentation:
    fonttheme: professionalfonts
---



## Install Some New Packages
```{r eval=FALSE}
install.packages('tree')
install.packages('randomForest')
library(MASS)
library(dplyr)
library(ggplot2)
library(raw)
library(tree)
library(randomForest)
```
```{r}
library(MASS)
library(dplyr)
library(ggplot2)
library(raw)
library(tree)
library(randomForest)
library(bindrcpp)
par(mfrow = c(1, 1))
```



## Get Some Data
```{r}
data(Boston)
head(Boston)
bos <- Boston %>% 
  dplyr::select(dis, rm, medv) %>% 
  mutate(medv_f = cut(medv, breaks = 0:5 * 10))
```



## Plot the Data
```{r}
plot(bos)
```



## Plot the Data
```{r}
plot(medv ~ rm, data = bos)
plot(medv ~ log(dis), data = bos)
```



## Maybe use ggplot2
```{r}
p <- ggplot(data = bos) +
  geom_point(mapping = aes(
    x = rm, 
    y = log(dis),
    color = medv
  )) +
  scale_color_gradientn(colours = c('yellow', 'green', 'blue', 'red'))
```



## Plot
```{r}
p
```



## Create a Decision Tree
```{r}
tbos <- tree(
  formula = medv ~ rm + log(dis),
  data = bos
)
summary(tbos)
```



## Plot the decision tree
```{r}
plot(tbos)
```



## Try Again
```{r}
plot(tbos)
text(tbos)
```



## Try Again #2
```{r}
plot(tbos, type = 'uniform')
text(tbos, pretty = 5, col = 'blue', cex = 0.8)
```



## Open the Region Experience data
```{r}
data(RegionExperience)
RegionExperience %>% head
```



## Take a Quick Look
```{r}
plot(RegionExperience)
unique(RegionExperience$Region)
unique(RegionExperience$PolicyYear)
```



## Let's Add a Frequency Field
```{r}
rex <- RegionExperience %>% 
  mutate(Freq = NumClaims / NumPolicies) %>% 
  dplyr::select(Region, PolicyYear, Freq)
unique(rex$Region)
```



## Run a Quick Tree
```{r}
trex <- tree(
  formula = Freq ~ .,
  data = rex
)
```



## Plot the tree
```{r}
plot(trex, type = 'uniform')
text(trex, pretty = 10, col = 'blue', cex = 0.8)
```



## Now, you try!
* Go back to the `Boston` dataset in the `MASS` package
* Pick some variables to model house value against
* Plot something
* Create a tree and plot it



## A sample
* crim = per capita crime by town
* ptratio = parent teacher ratio
* age = proportion of owner-occupied units built prior to 1940
```{r}
bos <- Boston %>% 
  mutate(medv_pr = medv / rm) %>% 
  dplyr::select(age, dis, medv_pr) %>% 
  mutate(medv_prf = cut(medv_pr, breaks = 0:5 * 2))
```



## Maybe use ggplot2
```{r}
p <- ggplot(data = bos) +
  geom_point(mapping = aes(
    x = age, 
    y = dis,
    color = medv_pr
  )) +
  scale_color_gradientn(colours = c('yellow', 'green', 'blue', 'red'))
```


## Plot
```{r}
p
```



## What does the tree look like?
```{r}
tbos <- tree(
  formula = medv_pr ~ age + log(dis),
  data = bos
)
```



## Plot It
```{r}
plot(tbos, type = 'uniform')
text(tbos, pretty = 5, col = 'blue', cex = 0.8)
```



## Random Forests
* What if you picked a random sample of rows BEFORE creating a tree?
* Then what if you took n random samples?
* Average the predictions of each tree
* Or, count up how often different characteristics are used as the FIRST break



## This Stuff is Advanced
* use the `randomForest` package
* How you parameterize the call is important, you can easily run out of memory



## Let's Try One
```{r}
rf <- randomForest(
  medv ~ ., data = Boston, 
  ntree = 100,
  nodesize = 100,
  importance = TRUE,
  mtry = 3)
rf
```



## Now what?
* `plot`
* `importance`
* `varImpPlot`
* `predict`



## `plot`
```{r}
plot(rf)
```


## `importance`
```{r}
importance(rf)
```


## `varImpPlot`
```{r}
par(mfrow = c(1, 2))
varImpPlot(rf)
```


