---
title: "Final Review"
author: "Brian A. Fannin"
date: "August 24, 2017"
output:
  revealjs::revealjs_presentation:
    center: no
    css: ./css/revealOpts.css
    reveal_plugins:
    - notes
    - zoom
    self_contained: no
    theme: solarized
    transition: slide
  tufte::tufte_handout: default
---

```{r include=FALSE}
knitr::opts_chunk$set(
    warning=FALSE
  , error=TRUE
  , echo=TRUE
  , message=TRUE
  , collapse=TRUE
  , fig.height = 4
)
knitr::opts_knit$set(root.dir = normalizePath('../../'))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
```

# Fix this code

## Q: Where's my data?

```{r}
data(MultiTri)
```

## A: 

```{r}
library(raw)
data(MultiTri)
```

## Q: But I just wanted the first two rows!

```{r}
thirdRow <- MultiTri[1:2]
```

## A: 

```{r}
thirdRow <- MultiTri[1:2, ]
```

## Q: What unused argument?

```{r}
devYear1997 <- MultiTri[DevelopmentYear = 1997, ]
```

## Q: But I used '==', not '=' !!

```{r}
devYear1997 <- MultiTri[DevelopmentYear == 1997, ]
```

## A: 

```{r}
devYear1997 <- MultiTri[MultiTri$DevelopmentYear == 1997, ]
```

## Q: 

See script window

## A: 

```{r}
dfStuff <- MultiTri %>% 
  select(Company)
```

## Q: But I thought "<-" assigned stuff?

```{r}
dfStuff <- MultiTri %>% 
  summarise(TotalPremium <- sum(NetEP))
```

## A:

```{r}
dfStuff <- MultiTri %>% 
  summarise(TotalPremium = sum(NetEP))
```

## dplyr sorbet

```{r}
dfLosses <- MultiTri %>% 
  filter(Company == "New Jersey Manufacturers Grp"
         , AccidentYear + Lag - 1 <= 1997
         , Line == "Workers Comp") %>% 
  select(-Line, -Company, -GroupCode, -Reserve1997, -Single, -DirectEP, -CededEP) %>% 
  arrange(AccidentYear, DevelopmentYear) %>% 
  group_by(AccidentYear) %>% 
  mutate(PriorIncurred = dplyr::lag(CumulativeIncurred)
         , PriorPaid = dplyr::lag(CumulativePaid)
         , IncrementalPaid = CumulativePaid - PriorPaid
         , IncrementalIncurred = CumulativeIncurred - PriorIncurred)
```

## Q: Why doesn't anything plot?

```{r}
library(ggplot2)
plt <- ggplot(dfLosses, aes(Lag, CumulativePaid))
plt
```

## A/Q: That's better, but it doesn't make any sense 

```{r}
plt <- ggplot(dfLosses, aes(Lag, CumulativePaid)) + geom_point()
plt
```

## A/Q: Maybe lines?

```{r}
plt <- ggplot(dfLosses, aes(Lag, CumulativePaid)) + geom_line()
plt
```

## A/Q: Maybe color?

```{r}
plt <- ggplot(dfLosses, aes(Lag, CumulativePaid, color = AccidentYear)) + geom_line()
plt
```

## A:

```{r}
plt <- ggplot(dfLosses, aes(Lag, CumulativePaid, color = as.factor(AccidentYear))) + geom_line() + geom_point()
plt
```

## Q: But you told me to use "=" in dplyr!

```{r}
dfLag2 <- dfLosses %>% 
  filter(Lag = 2)
```

## A:

```{r}
dfLag2 <- dfLosses %>% 
  filter(Lag == 2)
```

# Actionable takeaways

##

* Who knows how to cook?
* What's the difference between bread and beer?
* The unified theory of baking and Michael Ruhlman

## The R equivalent of Ratio

* The data frame is the basic unit of data organization
* A plot is the basic unit of visualization
* Linear fits are the most populare basic units of statistical models
* Markdown is the basic unit of sharing your research with other humans

## Stuff you can now do in R

* Fit a loss distribution
* Join policy and claim data
* Do basic reserve analysis
* Store your analysis __AND__ documentation in a Word file