---
title: "Visualization"
author: "R LAS"
date: "December 12, 2016"
output: slidy_presentation
---

# R plots


R plot has different methods depending on what the class of the object is. 

```{r}

methods(plot)
```


# A Simple Plots

```{r}
plot(x=1,y=1)
```

Features include:

- Box around plot area
- Point is represented by a circle
- No title or subtitle
- Default axis labels
- Axis limits are determined by data

# Saving Plots

You can save to a pdf, png, and other formats.
```{r}
pdf(file="samplescatterplot.pdf")
plot(x=runif(100),y=runif(100))
dev.off()

getwd()

```

# Plot Options

```{r}
p1<-par()
dev.off() #close the default device

class(p1)
length(p1)

p1[40]
```

# Plot Layout



`layout` divides the device up into as many rows and columns as there 
are in `matrix`, with the column-widths and the row-heights specified in the respective
arguments.

```{r}
layout(matrix(1:4,nrow=2))
layout.show(4)
```

# Plot Layout

To get a layout, create a matrix representing what you want.

```{r}
m1<-c(1,2,3,3,3,3)
m1<-
matrix(m1,nrow=2,
  byrow=FALSE)
m1
```


# Plot Layout

```{r}
layout(m1)
layout.show(3)
```


# Plot Layout

What commands give the following layout?

```{r, echo=FALSE}
m1<-matrix(c(1,1,1,2,3,3,3,4,3,3,3,4,3,3,3,4),
           nrow=4,byrow=TRUE)
layout(m1)
layout.show(4)
```

# Plot Layout
Solution.

```{r}
m1<-matrix(c(1,1,1,2,3,3,3,4,3,3,3,4,3,3,3,4),
           nrow=4,byrow=TRUE)
layout(m1)
layout.show(4)
```

# Plot Layout

You can create your own summary plot for exploratory data analysis.

```{r}
m1<-matrix(1:4,2,2)
x1<-rnorm(100)
layout(m1)

plot(x1)
hist(x1)
boxplot(x1)
qqnorm(x1)
```

# Plot Layout

Let's write this as a function.

```{r}
mysummaryPlot<-function(x){
  m1<-matrix(1:4,2)
	layout(m1)
	plot(x)
	hist(x)
	boxplot(x)
	qqnorm(x)
	}
```


# Plot Layout

```{r}

mysummaryPlot(x1)
```

# Colors
There's a default palette and a large number of colors available.

```{r}
palette()
length(colors())
head(colors())
```


# Colors

```{r}
set.seed(6345789)
numColors<-10
n1<-sample(1:length(colors()),replace=FALSE,
           size=numColors)
pie(rep(1,numColors),col=colors()[n1],
    labels=colors()[n1])
title(main="A Color Wheel")
```

# Colors

```{r}
set.seed(6345789)
numColors<-10
n1<-sample(1:length(colors()),replace=FALSE,
           size=numColors)
pie(rep(1,numColors),col=colors()[n1],
    labels=colors()[n1])
title(main="A Color Wheel")
```


# Colors

There are predefined palettes.

- rainbow
- heat.colors
- terrain.colors
- topo.colors
- cm.colors


```{r}
numColors<-8
pie(rep(1,numColors),col=rainbow(numColors),
    labels=rainbow(numColors))
title(main=paste("rainbow(",numColors,")",sep=""))
title(sub="Note that the names are in RGB")
```




# Colors

A very detailed discussion of R colors is available at <http://research.stowers-institute.org/efg/R/Color/Chart/>

The RColorBrewer package "creates nice looking palettes especially for thematic maps."
<http://colorbrewer2.org/>

```{r}
library(RColorBrewer)
display.brewer.pal(6,"RdBu")
```


# pch

```{r}
X1<-rep(1:5,5)
Y1<-c(rep(1,5),rep(2,5),rep(3,5),rep(4,5),rep(5,5))
plot(x=X1,y=Y1,pch=1:25,col="blue",bg="red",cex=2,bty="n", xlim=c(0.5,5), 
     ann=FALSE,axes=FALSE)
text(x=X1-.25,y=Y1,1:25)
title(main="Plotting Characters pch=1:25")
```

# bty

```{r}
btyopts<-c("n","l","7","c","u","]")
m1<-matrix(1:6,nc=2)
layout(m1)
for (i in 1:6){
plot(x=1,y=1,bty=btyopts[i])
  title(main=paste0('bty = ',btyopts[i]))}
```


# Titles and Axis Labels

`main` and `sub` refer to the main title and subtitle, respectively.
 Sometimes a plot will have a default title. If you want to supress that
you can use` main=NULL` and `sub=NULL`.
 `xlab` and `ylab` are the x-axis and y-axis labels, respectively.

```{r}
plot(x=1,y=1,xlab="Here's where xlab goes",
  ylab="Here's where ylab goes",type="n",bty="n")
title(main="Here's the main title",sub="Here's the subtitle")
```

# Axes Limits

```{r}
plot(x=1,y=1,
	xlim=c(-2,2),
	ylim=c(0,4)
)
```


# Text and Fonts

```{r}

plot(x=1,y=1,type="n",xlim=c(0,2),ylim=c(0,2),
     ann=FALSE,bty="n",
      xaxt="n",yaxt="n")
text(x=0,y=1.9,labels="Here's font=1 (default) ",
     font=0.5,adj=c(0,NA))
text(x=0,y=1.7,labels="Here's font=2 (bold)",
     font=2,adj=c(0,NA))
text(x=0,y=1.5,labels="Here's font=3 (italics)",
     font=3,adj=c(0,NA))
text(x=0,y=1.3,labels="Here's font=4 (bold italics)",
     font=4,adj=c(0,NA))
text(x=0,y=1.1,labels="Here's font=5 (symbols)",
     font=5,adj=c(0,NA))
```


# Magnification

```{r}
plot(x=1,y=1,type="n",xlim=c(1,2),ylim=c(1,2))
text(x=1.2,y=1.9,labels="Here's cex=.5",cex=.5)
text(x=1.2,y=1.7,labels="Here's cex=.75",cex=.75)
text(x=1.2,y=1.5,labels="Here's cex=1 (default)",cex=1)
text(x=1.2,y=1.3,labels="Here's cex=1.25",cex=1.25)
text(x=1.2,y=1.1,labels="Here's cex=1.5",cex=1.5)
```

# Line Types

- 0: blank
- 1: solid
- 2: dashed
- 3: dotted
- 4: dotdash
- 5: longdash
- 6: twodash

```{r}
plot(x=3,y=3,type="n",bty="n",
     ylab="lty",ylim=c(0,6),xaxt="n",xlab="",
     main="Line Type (lty=)")
abline(h=0:6,lty=0:6)
```

# legends
```{r}
plot(x=c(1,2),y=c(1,2),bty="n",col=c("red","blue"),pch=19)
legend("right",legend=c("Red Dot","Blue Dot"),
       fill=c("red","blue"))
```



# Bar Charts

```{r}
c1<-c(10,-2,5)
t1<-c("A","B","C")
col1<-c("blue","yellow","green")
barplot(c1,names.arg=t1,col=col1)
```

# Plotting Functions

There are two ways to plot functions. One is to plot the x and y values and connect the points. The other is to use `curve`.

```{r}
f1<-function(x) x*(x^2-1)
x1<-seq(from=-5,to=5,length=100)
y1<-f1(x1)

plot(x=x1,y=y1,type="l")
```

# Box and Whiskers

```{r}
library(datasets)
data(chickwts)
head(chickwts)
boxplot(weight~feed,
data=chickwts)
```



# Plotting Functions

```{r}
curve(f1,xlim=c(-5,5))
```



# A Sample Histogram

```{r}
t1<-rnorm(1000)
hist(t1)
```

Change some things:

- Default title
- Labels on x, y axes
- Subtitles
- Draw standard normal curve
- Change range of x-axis
- Include the equation of the curve

# A Sample Histogram

```{r}
main1<-'Histogram of 1,000 random  \n standard normal variables'

hist(t1,main=main1)
```

# A Sample Histogram

```{r}
sub1<-paste('Mean is', round(mean(t1),3),sep=' ')
hist(t1,main=main1,sub=sub1)
```

# A Sample Histogram

```{r}
hist(t1,main=main1,
	sub=sub1,
	freq=FALSE,
	xlab="Value", 
	ylab=
	"Probability",
	xlim=c(-5,5),
	ylim=c(0,.5))
```

# A Simple Histogram 

```{r}
hist(t1,main=main1,
	sub=sub1,
	freq=FALSE,
	xlab="Value", 
	ylab=
	"Probability",
	xlim=c(-5,5),
	ylim=c(0,.5))
curve(dnorm,col="red",add=TRUE)
```


# A Simple Histogram 

```{r}
hist(t1,main=main1,
	sub=sub1,
	freq=FALSE,
	xlab="Value", 
	ylab=
	"Probability",
	xlim=c(-5,5),
	ylim=c(0,.5))
curve(dnorm,col="red",add=TRUE)
text(x=3,y=0.4,
	expression(frac(1,sqrt(2*pi))* e^-frac(x^2,2)),
	cex=2)
```

# Chain Ladder Plot

```{r}
library(ChainLadder)
data(RAA)
class(RAA)
head(RAA)
```


# Chain Ladder Plot

```{r}
plot(RAA)
```


# Package tells R how to plot

```{r}
ChainLadder:::plot.triangle
```

# Class Project: Create This

```{r,echo=FALSE}
main2<-"Reinsurance Association of America \n Loss Development"
plot(RAA,pch=19,main=main2,xlab="Development Period (years)",bty="l",
	ylab="Loss Amount",xlim=c(1,11),cex.main=2,lty=1,
	xaxp=c(0,10,10))
legend("bottomright",legend=rownames(RAA),bty="n",fill=1:6,title="Accident Year")
```

# Class Project: Solution

```{r,echo=TRUE,eval=FALSE}
main2<-"Reinsurance Association of America \n Loss Development"
plot(RAA,pch=19,main=main2,xlab="Development Period (years)",bty="l",
	ylab="Loss Amount",xlim=c(1,11),cex.main=2,lty=1,
	xaxp=c(0,10,10))
legend("bottomright",legend=rownames(RAA),bty="n",fill=1:6,title="Accident Year")
