---
title: Loss Reserve Analysis
output: word_document
---

```{r include=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
knitr::opts_chunk$set(
  message = FALSE
  , warning = TRUE
  , error = TRUE
  , echo = FALSE
)
```

## Data Selection

1. From the NAIC data set, pick any company with at least three lines of business.

```{r}
library(raw)
data("comauto")
data("othliab")
data("wkcomp")
comauto$Line <- "Commercial Auto"
othliab$Line <- "Other Liability"
wkcomp$Line <- "Workers Compensation"
dfAll <- rbind(comauto, othliab, wkcomp)

dfCompanies <- dfAll %>% 
  group_by(Company) %>% 
  summarise(NumLines = dplyr::n_distinct(Line)) %>% 
  filter(NumLines == 3) %>% 
  select(-NumLines)

dfSome <- dfAll %>% 
  dplyr::inner_join(dfCompanies)

small_pay_cos <- dfSome %>% 
  filter(CumulativePaid <= 0) %>% 
  select(Company) %>% 
  dplyr::distinct() %>% 
  unlist()

dfSome <- dfSome %>% 
  filter(!Company %in% small_pay_cos)
```

```{r}
dfLosses <- dfSome %>% 
  filter(Company == "NC Farm Bureau Ins Grp") 
```

```{r}

```

## Forecast

2. Forecast all lines using at least the following techniques

    1. Calculate link ratios by writing your own algorithms

```{r}
dfLosses <- dfLosses %>% 
  arrange(Company, Line, AccidentYear, DevelopmentYear) %>% 
  group_by(Line, AccidentYear) %>% 
  mutate(PriorIncurred = dplyr::lag(CumulativeIncurred)
         , PriorPaid = dplyr::lag(CumulativePaid)
         , IncurredLDF = CumulativeIncurred / PriorIncurred
         , PaidLDF = CumulativePaid / PriorPaid)

dfUpper <- dfLosses %>% 
  filter(AccidentYear + Lag - 1 <= 1997)
```

```{r}
GetVectors <- function(df, which_lag, measure){
  
  df <- df %>% 
    filter(Lag == which_lag)
  
  prior <- df[[paste0("Prior", measure)]]
  current <- df[[paste0("Cumulative", measure)]]
  
}

LinkRatio <- function(current, prior, alpha = 0){
  
  ballast <- prior ^ (2 - alpha)
  ballast <- sum(ballast)

  weight <- (prior ^ (2 - alpha)) / ballast
  link <- current / prior
  factor <- sum(weight * link)

  factor
}
```

```{r}

LinkRatio(1:5, 6:10)
LinkRatio(1:5, 6:10, 1)
LinkRatio(1:5, 6:10, 2)

mojo <- dfLosses %>% 
  filter(Line == "Commercial Auto") %>% 
  LinkRatio(2, "Paid", 0)
```

    2. Determine ultimates using any of the techniques in the ChainLadder package
    3. Determine ultimates using a GLM, such as (overdispersed) Poisson
    4. Any other technique which you feel is appropriate

## Ranges

3. Develop ranges around the ultimates by accident year and in total

```{r}

```

